---
title: "Bike Share Analysis"
author: "Tılsım"
date: "2025-01-21"
output: html_document
---

```{r setup, include=FALSE}
# Set global chunk options
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This analysis explores bike-sharing data to uncover ride patterns and user behavior. 
The objective is to analyze how different user types (members vs. casual users) interact with the service, focusing on metrics such as ride duration, time-based patterns, and bike type preferences.

## Libraries and Data Setup

```{r}
# Load necessary libraries
library(tidyverse)
library(lubridate)
library(ggplot2)

# Load the dataset
bike_data <- read_csv("C:/Your/File/Directory/bikeshare_2023.csv")

# Display structure and summary of the dataset
str(bike_data)
summary(bike_data)
```

## Graph Settings

```{r}
# Remove scientific notation for better readability in visualizations
options(scipen = 999)
```

## Preprocessing the Data

### Ordering Weekdays for Consistency

```{r}
# Convert day names to ordered factors starting from Monday
bike_data$day_name <- factor(bike_data$day_name, 
                             levels = c("Monday", "Tuesday", "Wednesday", 
                                        "Thursday", "Friday", "Saturday", "Sunday"))
```

# Analysis and Visualizations

## Ride Patterns by User Type and Day of the Week

This section visualizes ride patterns across the week, segmented by user type.

```{r}
# Rides by user type and day of the week
ggplot(bike_data, aes(x = day_name, fill = member_casual)) + 
  geom_bar(position = "dodge") + 
  theme_minimal() + 
  labs(title = "Rides by User Type and Day of Week", 
       x = "Day of the Week", 
       y = "Number of Rides",
       fill = "User Type") +
  scale_fill_discrete(labels = c("Casual User", "Member"))
```

### Average Ride Duration by User Type

```{r}
# Average ride duration per day for each user type
bike_data %>%
  group_by(member_casual, day_name) %>%
  summarise(avg_duration = mean(ride_lenght)) %>%
  ggplot(aes(x = day_name, y = avg_duration, fill = member_casual)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Average Ride Duration by User Type",
       x = "Day of the Week",
       y = "Average Duration (minutes)",
       fill = "User Type") +
  scale_fill_discrete(labels = c("Casual User", "Member"))
```

## Hourly and Monthly Usage Patterns

### Adding Time-Based Columns

```{r}
# Create hour and month columns
bike_data <- bike_data %>%
  mutate(hour = hour(started_at),
         month = month(started_at))
```

### Hourly Ride Patterns

```{r}
# Hourly usage by user type, segmented by day
ggplot(bike_data, aes(x = hour, fill = member_casual)) +
  geom_bar(position = "dodge") +
  facet_wrap(~day_name, strip.position = "right") +
  theme_minimal() +
  labs(title = "Hourly Usage by User Type",
       x = "Hour of the Day (24-hour format)",
       y = "Number of Rides",
       fill = "User Type") +
  scale_fill_discrete(labels = c("Casual User", "Member")) +
  scale_x_continuous(breaks = seq(0, 23, by = 3))
```

### Monthly Ride Patterns

```{r}
# Convert month numbers to abbreviated names
bike_data$month_name <- month.abb[bike_data$month]
bike_data$month_name <- factor(bike_data$month_name, levels = month.abb)

# Monthly ride analysis
ggplot(bike_data, aes(x = month_name, fill = member_casual)) +
  geom_bar(position = "dodge") +
  theme_minimal() +
  labs(title = "Rides by User Type and Month",
       x = "Month",
       y = "Number of Rides",
       fill = "User Type")
```

## Bike Type Preferences

### Usage by User Type and Bike Type

```{r}
# Bike type usage by user type
ggplot(bike_data, aes(x = member_casual, fill = rideable_type)) +
  geom_bar(position = "dodge") +
  theme_minimal() +
  labs(title = "Bike Type Usage by User Type",
       x = "User Type",
       y = "Number of Rides",
       fill = "Bike Type") +
  scale_fill_discrete(labels = c("Classic Bike", "Docked Bike", "Electric Bike"))
```

### Percentage Analysis of Bike Type Usage

```{r}
# Calculate percentage of bike type usage
bike_data %>%
  group_by(member_casual, rideable_type) %>%
  summarise(count = n()) %>%
  group_by(member_casual) %>%
  mutate(percentage = count / sum(count) * 100) %>%
  arrange(member_casual, desc(percentage))
```

## Summarizing Data and Exporting Results

### Summary Table

```{r}
# Create summary statistics table
summary_table <- bike_data %>%
  group_by(member_casual) %>%
  summarise(
    total_rides = n(),
    avg_ride_length = mean(ride_lenght),
    median_ride_length = median(ride_lenght),
    max_ride_length = max(ride_lenght)
  )

# Export the summary table to a CSV file
write.csv(summary_table, "summary_bikeshare.csv", row.names = FALSE)